### 2017-10-09 response/response-jde-1.tex

\begin{center} \textbf{\large Income Shocks, Contraceptive Use, and
Timing of Fertility} \end{center}

\begin{center} Response to Referee Comments \end{center}

\noindent Please find attached the revised version of our paper,
``Income Shocks, Contraceptive Use, and Timing of Fertility.''
We would like to thank the two referees for the very useful comments and
suggestions,
which we believe has improved the paper.

[Additional discussion of overall things we have changed---if any]
- move discussion of education, wealth, and education results to appendix
to save space

@@

We rewrote the code for cleaning and analyzing the data from the ground up 
to make it easier to follow and eventually publish on GitHub.
This rewrite lowered the sample by three observations.
Including all married/partnered women up to age 45, rather than exclude
women younger than 18, the final sample size is, however, the same as 
before and the results are essentially unchanged.

We list below our responses to the individual comments and suggestions.

[Possible changes]
If the paper is too long, one possibility is to move the discussion of
age, education, and wealth descriptive statistics to the Appendix.
We only need this for the OLS and for the discussion of those factors,
which is now in the Appendix.

@@

It is correct that the current fixed effects model cannot tell
us whether we observe postponement of fertility or a permanent
reduction.
We focus on postponement for two reasons.

First, active postponement is a necessary---but not sufficient---condition
for permanent fertility reduction.
If we cannot show a conscious decision to delay fertility, it is 
hard to argue that we can ever find evidence of conscious permanent 
reduction in fertility.

Second, it is not clear that shocks---in and off themselves---should
substantially affect completed fertility. 
We discuss evidence from Guatemala that the overall \emph{risk} of 
experiencing a shock is likely to affect fertility, but that only
shocks that occur late in a woman's reproductive life cycle is
likely to affect completed fertility (because the opportunities to
make up for the ``missed'' birth are fewer and because the marginal
benefit of a child is smaller the more children you already have and
the older you are).
Hence, an important distinction is that long term fertility is likely
affected by the overall \emph{risk}, whereas the exact timing of
fertility is driven by a combination of risk and shocks (probably
mostly shocks).

The longitudinal nature of the data is important for our ability
to control for unobservable village, household, and individual 
characteristics, but the trade-off is that the data only cover 
32 months and that is too short to analyze long-term fertility
outcomes.
We followed your suggestion of adding an additional lag to 
the model and discuss the results in a new section in the Appendix.
There are, however, a number of issues with this approach.
First, it is possible that the effect of a crop loss is felt
in subsequent period and not only in the period in which it
was observed.
Second, we lose 1/4 of observations each time we include an 
additional lag, which leaves us with less variation because
the first round had a higher proportion of crop losses than
the subsequent rounds

Furthermore, even if shocks only had an impact in one period, 
there are two competing factors at work here.
First, a negative shocks will make people more likely to postpone 
their pregnancy/birth because of the factors discussed in the
paper.
Second, longer spacing between births can be costly because economies
of scale are harder to realize and because the mother may have to 
be out of the labor force for longer if she already have young
children to care for.
We have therefore also added an interaction between crop loss for 
different periods with the idea that there is a cost to continually 
postponing births.

If crop losses last more than one period and there is a cost 
associated with further postponement, the effect of both shocks 
on their own should be same and the interaction the opposite. 
Broadly speaking that appears to hold. 
The upshot---as far as we can tell---is that what we observe is 
postponement and not an overall limitation in fertility. 
This makes sense given results for Guatemala, where long-term risk 
affects the desire fertility \emph{level}, whereas shocks experienced 
affect the \emph{timing} of fertility. 
We have added a discussion of this in the Conclusion.

@@

[RESPONSE:]

We assume that the reference to ``modern'' is a typo and that the
first sentence is meant to refer to traditional contraceptives.
To be clear, we argue that both modern and traditional contraceptive
use is intentional.
It is correct that we cannot directly rule out that some women will 
report abstinence as a contraceptive strategy even if the abstinence 
arise from something other than a desire to limit the risk of pregnancy.
It is important, however, to note that the question was specifically
about contraception.

\begin{quote}
Some couples use contraception methods to avoid pregnancy or to
space births. Are you currently using a method of contraception?
(For example, the pill, the IUD, condom, withdrawal, rhythm,
abstinence.)
\end{quote}

Although we cannot measure directly measure stress, we do have information 
on illnesses and their durations.
If stress is an important factor in reporting abstinence, we should
expect the same to be the case with illness, especially those that 
prevent a woman from working.
Hence, although we did not explicitly cover this possibility in the original 
version, coital frequency may also decrease because of illness.
We have added that possibility to the text.

% We would expect illness to go up both because of the direct effect
% on the household and from increased stress.

We show the effect of crop loss on illness in Table 5.
Crop loss is  not associated with substantial enough changes in BMI or 
self-reported illness for the women in the sample to explain our fertility results.
Furthermore, we ran regressions (not included in the paper) that show that there 
is only a very weak positive association between having been ill for two or more 
days in the last 6 months and the likelihood of reporting using abstinence 
as a contraceptive strategy.
This relationship becomes only marginally stronger, and never anywhere near
statistically significant, if defining illness as not being able to
perform usual activities for 7 days or more.
Going above 7 days makes the relationship negative, most likely because
very few women report being unable to perform regular activities for that
long.

To further investigate the role of abstinence we added a new
table to the Appendix, which further subdivides the results in Table 6.
Table A-13 [TK] show, not surprisingly, that abstinence is the most 
effective traditional contraception, although only in combination 
with a crop loss as in the table in the main text.
A major caveat is that only 19 women report using abstinence only
across the three waves used for these regressions.
Modern contraceptive also are mainly effective in combination with
a crop loss, and the direct effect of using modern contraceptives
is not statistically significant.
We discuss these results at the end of the section on 
``Conscious Decision or Unintended Consequences?''.

@@

We may return to this question in a separate paper.
A problem with the current literature on timing of marriage
is that it tend to focus on only the female side of the marriage market.
A more satisfactory model will have to consistently explain both sides 
of the market.
Hence, results that show, for example, that women are less likely to
marry after a crop loss would have to show the same for men.

\item The survey measures crop loss continuously, and the indicator for
crop loss greater than 200 TZS is the authors' creation. They show in
the appendix that they obtain similar but less significant results if
they use the continuous measure (in levels or logs) or an indicator for
any crop loss. The 200 TZS seems arbitrary and possibly ex post, since
it leads to the most attractive results of the three sets of results
provided. It would be useful to more transparently show readers the
underlying variation and possible non-linearities, perhaps with a binned
version of the variable.

[RESPONSE:]

Our original discussion of this issue was scattered throughout the
original version of the paper.
We have now combined and expanded the discussion in the
Robustness section.
Before we respond, it is worth pointing out why we present the dummy
cut-off results in the paper.
We presented this paper a number of times before submission and whether 
we led with our dummy results or with the log results the discussant 
invariably asked us to focus on the other set of results.
We settled on the dummy results because they are easier to interpret 
for most people, but the downside is, as you point out, that the 
cut-off used is subject to discussion and that there is not one right level.

In many ways, the log specification in Table A- TK is more theoretically 
appealing because it captures that we should expect a given increase
in crop loss to have a diminishing impact on fertility related outcomes 
as the size of the crop loss increase (which is also why the linear version is 
not appealing; it assumes that a one TZS increase in crop loss has
the same effect whether it is at 0 or 50,000 TZS). 
Furthermore, the conclusion one would draw from the log results
is exactly the same as what we draw from the dummy approach.
Crop losses lead to statistically significant reductions in the likelihoods 
of pregnancy and birth with crop losses, and statistically significant 
increases in the likelihood of using contraceptives overall and
for the use of traditional contraceptives.
The differences from the dummy results are that the pregnancy and
birth results are slightly less statistically significant, whereas
the contraceptives results are slightly more statistically significant.
The main downside is that the log version is more difficult to interpret.

To show that the specific cut-off chosen does not have substantial effects on
the conclusions drawn we have added Table A-TK, which shows the main results
for different cut-offs together with the associated percentages of observations
that would be counted as having a crop loss.
Except for any crop loss (cut-off of above zero) the results
are consistent across all other specifications.
As the cut-off becomes very large statistically significance is lower
for pregnancy and births and eventually also for contraceptive use.
This is not surprising.
As we increase the cut-off there are more and more households 
that have experienced a crop loss but are not counted as having a
crop loss, which will tend to bias results towards zero if there
is an effect of crop loss for levels lower than the cut-off 
(this is confirmed by regressions---not
shown---where we drop all observations with crop loss larger than zero but
smaller or equal to the cut-off and find substantially larger coefficients).

@@

Even with family fixed effects, crop loss is potentially tied with
omitted variables. Crop loss is subjectively measured, so we'd need to
rely on the claim of the survey respondent that the loss was for
"uncontrollable" circumstances. The use of family fixed effects does
address fixed differences in wealth or education that might be related
to families at greater risk of crop loss. But, there could still be time
varying factors. We can't be certain that the crop loss is not the
result of some disease epidemic (or conflict) that also could affect
fertility.

Suggestion: The manuscript should be a little more forthcoming about
this potential problem. I would like to see an exploration into the
potential mechanisms behind the crop loss. Is it weather?

[RESPONSE:]

[see response to R1's q7 on time-varying covariates]

[TK we need to tone down the causality part; explain the growing season and
the crop loss question]

% The question ask (Question 11.B.9): ``Have you lost any part of the 
% harvested crop to insects, rodents, fire, rotting, etc.?''
It is correct that the crop loss is subjectively measured and that
it is not possible to verify what factors played a role in the crop loss.
As you suggest, weather is a potential factor.
Unfortunately the only available weather data consist of monthly rainfall 
at district level (there are five districts in the data) with missing 
information for some months.
Furthermore, there is no 30 year average to compare the observed 
district level rainfall data to.
This means that it is not possible to reliably separate out positive 
(above mean) and negative (below mean) rainfall shocks.
Although community or women fixed effects will partly ameliorate the
problem of missing average data, we cannot rule out the possibility
that all observed periods deviated substantially from the average in 
the same direction in which case we will fail to capture that the observed
rainfalls were all ``shocks''.

@@

A number of different specifications of rainfall's association with 
crop loss is possible.
Kagera has two growing seasons, February through May and
October through December \citep[p 6]{Tanzania2007}.
We tried two different strategies: 
one where rainfall is measured as the average monthly rainfall 
over a set period prior to the survey month; 
and one based on average rainfall for growing season(s) preceding 
the survey month 
(that is, not including the current growing season if the survey
took place during a growing season).
Despite that rainfall is measured at district level either strategy
lead to variation across communities because of differences in 
when each community was surveyed.

The first strategy did not provide any clear results for a variety
of period lengths and we therefore focus on the growing season approach.
Table TK shows individual level fixed effects estimates of the 
effects of average monthly rainfall in centimeters on the likelihood
of experiencing a crop loss.
We use two specification, one with the last prior growing season and 
one with prior growing for waves 2 through 4 and the last two prior 
growing seasons for wave 1.
Higher rainfall is associated with a lower likelihood of a crop loss. 
A one standard deviation change in rainfall is associated with an
approximately 6 percentage point change in the probability of 
experiencing a crop loss.
We also tried adding square terms, but the estimated coefficients
were small and far from statistically significant.
The same is the case for using crop loss per capita as the dependent
variable.

It is worthwhile noting the relatively small difference in the 
estimated effect of rainfall between the two models, 
although the coefficient on rainfall in Model I is just outside
the standard levels of statistically significance.
This suggests that the high level of crop loss observed in
wave 1 (see discussion below as well) is not mainly due to 
rainfall shocks that occurred longer ago than for the other waves.

In both models we include a dummy for missing information on rainfall 
for any month in the period / district combination.
If the rain data has one or more missing values for the months
covered in the growing season there is a large and statistically 
significant increase in the probability of reporting a crop loss
(dropping observations with missing information does little
to change the coefficients on rainfall, but do make them
statistically insignificant).
It is, however, very difficult to establish why this is because
no additional information in provided with the data.
Furthermore, the missing information problem only affects
observations in waves 2 and 3.

@@

We also tried including rainfall---as defined above---as 
additional variables in the main regressions.
Consistent with the small effects of rainfall on crop loss
there are no apparent effects of the rainfall measures on the
outcomes and little to no change in the coefficients for crop
loss on our five main outcomes.

We realize that this answer does fully explain where the
crop losses come from but unfortunately there is not much
more relevant information that we can find beside the
rainfall data and the variables discussed in our response
to Reviewer 1.

@@

I am concerned by the fact that crop loss is much higher in the first
and second waves of the survey (see Table 1 on page 6). So, there's a
chance that the estimates are driven by some time trend that affects
family types that are more likely to experience crop losses. (The survey
wave fixed effects won't deal with this if there are differential trends
in the type of families who experience crop loss.)

Suggestion: Control for family characteristics (age, proxies for
education, proxies for wealth) at the first wave interacted with survey
wave. At the very least, this will help acknowledge the identification
threat for the reader.

[RESPONSE:]

It is, indeed, potentially concerning that there are so substantial
differences in the proportion of households that experienced a 
shocks over the different rounds, although it is possible that
this simply is a stark reminder of the volatile nature of life
in a developing country.

We have added an additional table to the Appendix to address this
concern and discuss the results in the Robustness section. 
In addition to your suggestion of interacting a linear
time trend with wave 1 characteristics (assets per capita,
age group dummies, and education group dummies), we also present
a separate model where wave dummies are interacted with the
same set of wave 1 characteristics.
The small differences in point estimate and level of significance 
indicate that our results are not driven by differential trends
in which types of households that experience crop loss.

@@

Neighbors may be affected an individuals crop loss. For one, price of
food goes up. This could be a positive income shock for families with
crops, but bad for families working in other areas.

Suggestion: Aggregate crop loss to the community level and use that as
the treatment variable.

[RESPONSE:]

We already include a version of this in the paper.
Table A-17 TK include the fraction of households in the community 
who have experienced a crop loss (excluding the household itself).
The sign of the village level crop loss are generally the same as
those for the household crop loss, although---with the exception
of traditional contraceptives---the village level variable is not
statistically significant.
Furthermore, the crop loss coefficients decline only slightly
when including village level crop loss indicating that there is
only relatively mild spill-over between households for this
particular type of shock.

@@

\item Births analysis

The births analysis automatically drops first wave of survey due to the
longer lag in reported crop loss (Table 3). However, I did not see this
mentioned anywhere.

[RESPONSE:]

The discussion of the longer lag was hidden in the middle of the 
Estimation Strategy section and the motivation was not clear.
We have rewritten that part of the Estimation Strategy to better
explain why the lag is there.

### 2017-10-09 paper/incomeShocks-jde-r1.tex
Edits of paper and reviewer response

\usepackage[margin=1.0in]{geometry}
\usepackage[figuresleft]{rotating}
\usepackage[longnamesfirst]{natbib}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[flushleft]{threeparttable}
\usepackage{setspace}
\usepackage[justification=centering]{caption}
\usepackage[font=scriptsize]{subfig}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\usepackage{tikz}
\usetikzlibrary{decorations}
\usepackage{xfrac}

% Checking for unused references/labels - comment out before final run
\usepackage{refcheck}

@@

\title{Income Shocks, Contraceptive Use,\\ and Timing of Fertility\thanks{%
We would like to thank 
Alfredo Burlando,
Zo\"e McLaren, 
Erik Plug,
Lawrence Wu, 
and seminar participants at 
Agricultural Economics Society Annual Conference,
European Society for Population Economics Annual Conference,
IUSSP International Population Conference,
Montana State University, 
NEUDC 2015,
Norwegian School of Economics, 
Pacific Conference for Development Economics, 
the seventh annual PopPov Conference, 
Population Reference Bureau,
University of Oregon,
University of Washington,
and the 
Western Economic Association International Conference
for helpful comments and suggestions.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}}

@@

\begin{abstract} 
% JDE abstract
\noindent This paper examines the relationship between income 
shocks and fertility decisions. 
Using panel data from Tanzania, we estimate the impact of agricultural shocks 
on 
pregnancies, 
births, 
and 
contraception use.
The likelihood of 
pregnancies and childbirth are significantly lower for households that experience
a crop shock.
Furthermore, women significantly increase their contraception use in response 
to crop losses. 
The increase in contraceptive use comes almost entirely from traditional 
contraceptive methods, such as abstinence, withdrawal, and the rhythm method.
We argue that these changes in behavior are the result of deliberate decisions of
the households rather than the shocks' effects on other factors that influence
fertility, such as women's health status, the absence or migration of a spouse, 
the dissolution of partnerships, or the number of hours worked. 
We also show that, although traditional contraceptives have low overall
efficacy, households with a strong incentive to postpone fertility
are very effective at using them.

@@

How shocks affect household decisions on fertility in developing countries has 
received little attention in the literature.
A correlation between shocks and the timing of births is well-established, but it is 
unclear whether the cause is households' intentional planning or unintended consequences 
of the shocks \citep{pitt98b,lindstrom99,Evans2010,Portner2014}.%
\footnote{
There is a related literature on the effects of weather, especially temperature,
on timing of fertility \citep[see, for example,][]{Barreca2015}.
%  and TK Josh paper.
In a different vein, a monthlong blackout in Zanzibar increased 
births 8 to 10 months later \citep{Burlando2014}.
There is also a large literature examining how households in developing countries
cope with shocks in areas such as labor supply, crop choice, and education.
}
There is, in general, little evidence on what determines the timing 
of fertility in developing countries, with most of the literature instead 
focusing on the number of children born \citep{Portner2018}.
% [Little evidence in general on timing of fertility in developing countries;
% most of the research has focused on completed fertility / children-ever-born] 
% Furthermore, standard models of fertility focus mainly on the effects of time
% invariant household and individual characteristics on fertility outcomes, and 
% tend to ignore the dynamic aspects of fertility decisions.%
% \footnote{
% For an early exception see \cite{Newman1984}.
% }

@@

Most of the prior literature on the impact of economic and other shocks on fertility 
has used historic data on what are now developed countries.
Data from Rouen, France, over the period 1681-1787, show that increases in the
price of wheat led to a dramatic fertility decline for the urban poor, 
whereas fertility of the urban wealthy was unaffected \citep{Galloway1987}.
Similarly, English data from 1542 to 1800 show that mortality shocks led to short-term
fertility declines, with the largest decline typically occurring  the year following 
the shock \citep{Bailey1998}.%
\footnote{
This data also indicated that an increase in real wage led to an increase in short-term 
fertility.
Increases in wage rate and declines in child mortality explain a significant part of the 
long-term fertility decline in Sweden during the period 1736 to 1946 \citep{Eckstein1999a}.
In another study on Sweden, increases in the value of women's time relative to men's 
time---arising from increases in the price of butter---led to a decline in fertility during 
the  1860-1910 period \citep{Schultz1985}. 
}
In pre-industrial Sweden, positive crop shock, positive weather shock, and 
positive wage shock all increased fertility for a five- to ten-year period following 
the shocks \citep{Eckstein1985}.
However, the increase in birth rate changed only the timing of births with no 
cumulative effect on the long-term fertility rate.
Similarly, the increasing price of rye in Southern Sweden led landless and 
semi-landless families to reduce their fertility, whereas landed farmers 
showed no change \citep{bengtsson06}.

Using present-day data, there is evidence that famine and domestic/regional 
military attacks in Ethiopia led to short-term decreases in the likelihood of 
conception \citep{lindstrom99}.
Closer to our paper, positive rainfall shocks in Senegal increased 
fertility \citep{pitt98b}.
None of these paper can, however, directly address whether the observed changes 
arise from deliberate control of fertility.

@@

The paper makes three main contributions to the literature. 
First, this is the first paper to show that households respond to income shocks 
through contraceptive use, which, in turn, affects the timing of child bearing. 
% Our study provides evidence that the postponement of fertility occurs through a 
% planned decision process rather than as an unplanned consequence.
Second, with low rates of modern contraceptive use in the area, the paper contributes 
to the literature on how populations control fertility before widespread adoption of
modern contraception.%
\footnote{
There has long been a discussion of when and how families 
undertook fertility control before the emergence of modern 
contraceptives \citep{Guinnane2011,Cinnirella2017}.
Simulations show that it clearly is possible to substantially limit fertility
even with only access to traditional contraceptives, but there is only 
limited evidence available on exactly which methods were used 
\citep{Michael1976,David1986,Santow1995}.
}
Finally, the paper contributes to the consumption smoothing literature, as households 
affected by income shock use family planning as a mechanism to smooth their consumption. 
In the short run, bearing a child means diverting scarce resources away from other
useful purposes to the birth and maintenance of the child. 
Farmers can, therefore, partly smooth their consumption by delaying births during 
times of income shock.

@@

The data contain information on individual and household level demographic 
and socio-economic characteristics. 
In addition, questions about fertility and birth control were asked of all 
married women, regardless of age, and unmarried women 14 years and older. 
We restrict our sample to married or partnered women 18-45 years of age who we 
observe in all 4 rounds of the survey.
This ensures that we capture women who are at risk of conception and for whom 
the available observations cover sufficient time for shocks to have an effect
on pregnancies and births.
The downside is that the sample size is reduced.
We discuss possible non-random attrition below.

@@

Table \ref{tab:desc_stat_croploss} shows the main explanatory variable of interest 
and the 5 outcomes that we examine.
The main explanatory variable is a dummy for whether the household experienced 
an accidental loss of crops, which covers any crop lost due to insects, 
rodents, fire, rotting, or other calamities, greater than TZS 200 since the 
prior survey round (or prior 1 year for the first round).
% We create a dummy for accidental crop losses greater than 200 TZS,
% which is our preferred measure of income shock.
The TZS 200 cut-off is equivalent to approximately 5th 
percentile of all household savings.
% That period is approximately 6 to 7 months for the last three surveys, but covers 
% the previous 12 months for the first survey. 
% [MAY HAVE TO EXPLAIN MORE ABOUT WHAT THIS MEANS FOR OUR ESTIMATES]
There is substantial variation in the number of households that experience a 
crop loss between the 4 survey rounds.
In the 12 months prior to the first survey round just over 70\% of households 
lost crops, whereas just over 25\% lost crops between the 1st and 2nd survey round.
That falls even further for the 3rd and 4th rounds, with less than 5\% 
reporting crop loss of more than TZS 200 for both.
Our results are robust to defining crop loss as any crop loss, with the 
coefficients slightly smaller and slightly less statistically 
significant as expected. 
% Crop losses between the valuation of 0-200 TZS are not coded as an income shock,
% because of the small magnitude (equal to  of savings). 
% and unlikely to affect fertility decisions of households. 
Furthermore, our results are also qualitatively similar if we use a log of the amount of 
crop lost or a linear specification of crop loss instead of the dummy variable.%
\footnote{
Results using these alternative specifications of crop loss are available in the Appendix.
}

On average across the four survey rounds, 17\% of women report being pregnant 
at the time of the interviews and 17\% of women report giving birth since the 
prior survey round.
In total, just below 70\% of women in the sample report giving birth in
the covered period (3\sfrac{1}{2} years).
Similarly, 70\% report that they are currently pregnant at one point during 
the 4 rounds.

We capture contraceptive use at the time of the survey, and further divide
contraceptives into traditional and modern types. 
Traditional contraceptives include abstinence, rhythm, and withdrawal, 
and modern contraceptives include condom, diaphragm, pill, IUD, injection,
and female and male sterilization.%
\footnote{
The relevant questions are "Some couples use contraception methods to avoid
pregnancy or to space births. Are you currently using a method of contraception?".
If the woman answers yes, she is then asked "What contraceptive method are you
and your partner using at present?"
}
A small subset of women report using both modern and traditional contraceptives;
these women are included in both categories.
If a woman is currently pregnant, we code her as not using contraceptives.
We remove sterilized women from the sample, as sterilization is typically a 
permanent procedure.

Contraceptive prevalence is low; at the time of the first survey round, 15\% 
of women report using any type of contraceptive---split between 11 percent that use 
traditional contraceptives and 4\% that use modern contraceptives.
In subsequent survey rounds, the contraceptive use rate hovers around 10\%, 
with 4 to 7\% using traditional contraceptives and 2 to 5\% 
using modern contraceptives.

@@

Table \ref{tab:desc_stat_women} shows descriptive statistics for basic
demographic and household characteristics at wave 1.
% This survey suffers from two data quality issues: misreporting of 
% age and misreporting of education level.
As the time span of the surveys is too short to confidently identify age effects,
we use women's age at the time of the first survey round.
There is substantial measurement error in age and heaping of reported ages 
around numbers ending in 0 and 5.%
\footnote{
Some women report the same age in all survey rounds despite that there are
21 months between the first and last survey round, whereas other women's reported 
age goes up by 1 year between each survey round despite that the surveys are only 
6 to 7 months apart.
}
% If her reported age vary more than the length of interval between each survey 
% round, we use her median reported age rounded down.
We use respondent's median reported age, rounded down, and 5-year age groups 
centered around ages ending in 0 and 5 to take account of heaping.
Only married/partnered women are in the sample, which explains the 
equal age distribution despite population growth in the area.

>>>>

\usepackage[margin=1.0in]{geometry}
\usepackage[figuresleft]{rotating}
\usepackage[longnamesfirst]{natbib}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[flushleft]{threeparttable}
\usepackage{setspace}
\usepackage[justification=centering]{caption}
\usepackage[font=scriptsize]{subfig}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\usepackage{xfrac}

% Checking for unused references/labels - comment out before final run
% \usepackage{refcheck}

@@

\title{Income Shocks, Contraceptive Use,\\ and Timing of Fertility\thanks{%
We would like to thank 
Alfredo Burlando,
Zo\"e McLaren, 
Erik Plug,
Lawrence Wu, 
two anonymous reviewers, 
and seminar participants at 
Agricultural Economics Society Annual Conference,
European Society for Population Economics Annual Conference,
IUSSP International Population Conference,
Montana State University, 
NEUDC 2015,
Norwegian School of Economics, 
Pacific Conference for Development Economics, 
the seventh annual PopPov Conference, 
Population Reference Bureau,
University of Oregon,
University of Washington,
and the 
Western Economic Association International Conference
for helpful comments and suggestions.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}}

@@

\begin{abstract} 
% JDE abstract
\noindent This paper examines the relationship between income 
shocks and fertility decisions. 
Using panel data from Tanzania, we estimate the impact of agricultural shocks 
on 
pregnancies, 
births, 
and 
contraception use.
The likelihood of 
pregnancies and childbirth are significantly lower for households that experience
a crop shock.
Furthermore, women significantly increase their contraception use in response 
to crop losses. 
The increase in contraceptive use comes almost entirely from traditional 
contraceptive methods, such as abstinence and the rhythm method.
We argue that these changes in behavior are the result of deliberate decisions of
the households rather than the shocks' effects on other factors that influence
fertility, such as women's health status, the absence or migration of a spouse, 
the dissolution of partnerships, or the number of hours worked. 
We also show that, although traditional contraceptives have low overall
efficacy, households with a strong incentive to postpone fertility
are very effective at using them.

@@

How shocks affect fertility decisions in developing countries has 
received little attention in the literature.
A correlation between shocks and the timing of births is well-established, but it is 
unclear whether the cause is households' intentional planning or unintended consequences 
of the shocks \citep{pitt98b,lindstrom99,Evans2010,Portner2014}.%
\footnote{
There is a related literature on the effects of weather, especially temperature,
on timing of fertility \citep[see, for example,][]{Barreca2015}.
%  and TK Josh paper.
In a different vein, a monthlong blackout in Zanzibar increased 
births 8 to 10 months later \citep{Burlando2014}.
There is also a large literature examining how households in developing countries
cope with shocks in areas such as labor supply, crop choice, and education.
}
Furthermore, there is, in general, little evidence on what determines timing 
of fertility in developing countries, with most of the literature instead 
focusing on the number of children born \citep{Portner2018}.
% [Little evidence in general on timing of fertility in developing countries;
% most of the research has focused on completed fertility / children-ever-born] 
% Furthermore, standard models of fertility focus mainly on the effects of time
% invariant household and individual characteristics on fertility outcomes, and 
% tend to ignore the dynamic aspects of fertility decisions.%
% \footnote{
% For an early exception see \cite{Newman1984}.
% }

@@

Most of the prior literature on the impact of economic and other shocks on fertility 
has used historic data on what are now developed countries.
Data from Rouen, France, over the period 1681-1787, show that increases in the
price of wheat led to a dramatic fertility decline for the urban poor, 
whereas fertility of the urban wealthy was unaffected \citep{Galloway1987}.
Similarly, English data from 1542 to 1800 show that mortality shocks led to short-term
fertility declines, with the largest decline typically occurring  the year following 
the shock \citep{Bailey1998}.%
\footnote{
This data also indicated that an increase in real wage led to an increase in short-term 
fertility.
Increases in wage rate and declines in child mortality explain a significant part of the 
long-term fertility decline in Sweden during the period 1736 to 1946 \citep{Eckstein1999a}.
In another study on Sweden, increases in the value of women's time relative to men's 
time---arising from increases in the price of butter---led to a decline in fertility during 
the  1860-1910 period \citep{Schultz1985}. 
}
In pre-industrial Sweden, positive crop shock, positive weather shock, and 
positive wage shock all increased fertility for a five- to ten-year period following 
the shocks \citep{Eckstein1985}.
However, the increase in birth rate changed only the timing of births with no 
cumulative effect on the long-term fertility rate.
Similarly, the increasing price of rye in Southern Sweden led landless and 
semi-landless families to reduce their fertility, whereas landed farmers 
showed no change \citep{bengtsson06}.
Using present-day data, there is evidence that famine and domestic/regional 
military attacks in Ethiopia led to short-term decreases in the likelihood of 
conception \citep{lindstrom99}.
Closer to our paper, positive rainfall shocks in Senegal increased 
fertility \citep{pitt98b}.
None of these paper can, however, directly address whether the observed changes 
arise from deliberate control of fertility.

@@

The paper makes three main contributions to the literature. 
First, this is the first paper to show that households respond to income shocks 
through contraceptive use, which, in turn, affects the timing of child bearing. 
% Our study provides evidence that the postponement of fertility occurs through a 
% planned decision process rather than as an unplanned consequence.
Second, with low rates of modern contraceptive use in the area, the paper contributes 
to the literature on how populations control fertility before widespread adoption of
modern contraception.%
\footnote{
There has long been a discussion of when and how families 
undertook fertility control before the emergence of modern 
contraceptives \citep{Guinnane2011,Cinnirella2017}.
Simulations show that it clearly is possible to substantially limit fertility
even with only access to traditional contraceptives, but there is only 
limited evidence available on exactly which methods were actually used 
\citep{Michael1976,David1986,Santow1995}.
}
Finally, the paper contributes to the consumption smoothing literature, as households 
affected by income shock use family planning as a mechanism to smooth their consumption. 
In the short run, bearing a child means diverting scarce resources away from other
useful purposes to the birth and maintenance of the child. 
Farmers can, therefore, partly smooth their consumption by delaying births during 
times of income shock.

@@

The data contain information on individual and household level demographic 
and socio-economic characteristics. 
In addition, questions about fertility and birth control were asked of all 
married women, regardless of age, and unmarried women 14 years and older. 
We restrict our sample to married or partnered women 17-45 years of age who we 
observe in all 4 rounds of the survey.
This ensures that we capture women who are at risk of conception and for whom 
the available observations cover sufficient time for shocks to have an effect
on pregnancies and births.
The downside is that the sample size is reduced.
We discuss possible non-random attrition below.

@@

Table \ref{tab:desc_stat_croploss} shows the main explanatory variable of interest 
and the 5 outcomes that we examine.
The main explanatory variable is a dummy for whether the household experienced 
an accidental loss of crops, which covers any crop lost due to insects, 
rodents, fire, rotting, or other calamities, greater than TZS 200 since the 
prior survey round (or prior 1 year for the first round).
There is substantial variation in the number of households that experience a 
crop loss between the 4 survey rounds.
In the 12 months prior to the first survey just below 70\% of households 
lost crops, whereas 25\% lost crops between the 1st and 2nd surveys.
That falls even further for the 3rd and 4th rounds, with less than 3\% 
reporting crop loss of more than TZS 200 for both.
The TZS 200 cut-off is equivalent to approximately 5th 
percentile of all household savings.
Our results are robust to alternative definitions of crop loss as we
discuss in Section \ref{sec:robustness}.

On average across the four surveys, 17\% of women report being pregnant 
at the time of the interviews and 16\% of women report giving birth since the 
prior survey round.
In total, just below 70\% of women in the sample report giving birth in
the covered period (approximately 3\sfrac{1}{2} years).
Similarly, 70\% report a pregnancy at one point during the 4 rounds.

We capture contraceptive use at the time of the survey, and further divide
contraceptives into traditional and modern types. 
Traditional contraceptives include abstinence, rhythm, and withdrawal, 
and modern contraceptives include condom, diaphragm, pill, IUD, injection,
and female and male sterilization.%
\footnote{
The relevant question is "Some couples use contraception methods to avoid
pregnancy or to space births. Are you currently using a method of contraception?"
If the woman answers yes, she is then asked "What contraceptive method are you
and your partner using at present?"
}
A small subset of women report using both modern and traditional contraceptives;
these women are included in both categories.
If a woman is currently pregnant, we code her as not using contraceptives.
We remove sterilized women from the sample, as sterilization is typically a 
permanent procedure.

Contraceptive prevalence is low; at the time of the first survey round, 14\% 
of women report using any type of contraceptive---split between 11 percent that use 
traditional contraceptives and 4\% that use modern contraceptives.
In subsequent survey rounds, the contraceptive use rate is just below 10\%, 
with 4 to 6\% using traditional contraceptives and 2 to 5\% 
using modern contraceptives.

@@

Table \ref{tab:desc_stat_women} shows descriptive statistics for basic
demographic and household characteristics at wave 1.
As the time span of the surveys is too short to confidently identify age effects,
we use women's age at the time of the first survey round.
There is substantial measurement error in age and heaping of reported ages 
around numbers ending in 0 and 5.%
\footnote{
Some women report the same age in all survey rounds despite that there are
21 months between the first and last survey round, whereas other women's reported 
age goes up by 1 year between each survey round despite that the surveys are only 
6 to 7 months apart.
}
We use respondent's median reported age, rounded down, and 5-year age groups 
centered around ages ending in 0 and 5 to take account of heaping.
Only married/partnered women are in the sample, which explains the 
equal age distribution despite population growth in the area.

A notable characteristic is the high level of reported education for a poor 
rural area, although there also appears to be errors in 
reported education levels.%
\footnote{
For many women, the reported education level decreases in certain survey 
rounds, or increases by 2 years or more in consecutive survey rounds, which 
conflicts with the 6 to 7 months survey intervals. 
Additionally, individuals are supposed to complete primary school by the age of 15,
so changes in primary schooling levels in our sample is likely from reporting 
errors. 
For women who report different primary school education levels, we 
assign the modal education level if possible, or, if there is no modal 
observation, the median years of education.
We use the same approach for household heads' education
when we use household heads' education as a proxy for wealth.
}
We divide education levels into three groups: no education, 
some primary (1-6 years of education), and primary school 
or above (7 or more years of education).
Only 29\% of the women received no education, 27\% received some 
primary education, and 44\% completed at least primary education.
The high reported education level is most likely the result of the 1974 
Universal Primary Education Movement,
which increased accessibility of primary education and enrollment rates, 
even though there are reports that the quality of education was very low.%
% In addition, the crisis Tanzania experienced in the 1980s further lowered the 
% quality and enrollments declined significantly.%
\footnote{
See \cite{Galabawa2001} and \cite{Wedgwood2005} for detailed discussions 
of the development of education in Tanzania.
}
Hence, it is unclear to what extent reported education levels reflect 
women's actual human capital.
